Efficiently Handling Guards in the Low Level Design of RPython's tracing JIT

### 2012-08-07 talk/vmil2012/paper.tex
add line numbers (puh)

Based on the informal observation that guards are among the most common
operations in the traces produced by PyPy's tracing JIT and that guards are
operations that are associated with an overhead to maintain information about
state to be able to rebuild the execution state in case of deoptimization, our
goal is to present concrete numbers for the frequency and the overhead produced
by guards, explain how they are implemented in the different levels of PyPy's
tracing JIT and explain the rationale behind the design decisions based on the
numbers.

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed, this process is described in more detail in
Section~\ref{sec:Resume Data}, during the recording phase special operations,
\texttt{guards}, are inserted into the recorded trace at all points where
control flow could diverge. As can be seen on Figure~\ref{fig:guard_percent}
guards account for 14.42\% to 22.32\% of the operations before and for 15.2\%
to 20.12\% of the operations after the optimization pass over the traced and
compiled parts of the benchmarks, making guards one of the most common
operations. Many of these guards fail rarely on not all during execution. Given
that associated with each guard information is stored, that is required to
rebuild the execution state in case control flow diverges from the recorded
path at a guard it is important to store the information associated with the
guards in a manner that tries to keep the overhead for storing the information
low while avoiding to put a burden on the execution of the recorded trace,
making the optimization of guards an important aspect of
the low-level design of a tracing just-in-time compiler.

@@

\todo{set line numbers to the line numbers of the rpython example}
\begin{figure}
    \input{figures/log.tex}
    \caption{Optimized trace}
    \label{fig:trace-log}
\end{figure}
% section Resume Data (end)
\section{Guards in the Backend}
\label{sec:Guards in the Backend}

>>>>

Based on the informal observation that guards are among the most common
operations in the traces produced by PyPy's tracing JIT and that guards are
operations that are associated with an overhead to maintain information about
the execution state to be able to rebuild it in case of deoptimization, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of PyPy's
tracing JIT and explain the rationale behind the design decisions based on the
numbers provided here.

The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed, this process is described in more detail in
Section~\ref{sec:Resume Data}, during the recording phase special operations,
referred to as \texttt{guards}, are inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for 14.42\% to 22.32\% of the
operations before and for 15.2\% to 20.12\% of the operations after the
optimization pass over the traced and later compiled parts of the benchmarks,
making guards one of the most common types of operations. Many of these guards
fail rarely or not all during execution. There are several aspects to consider
in the design and optimization of guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace,  - which is
the common case - should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the low-level design of a tracing just-in-time compiler.

@@

\begin{figure}
    \input{figures/log.tex}
    \caption{Optimized trace}
    \label{fig:trace-log}
\end{figure}
% section Resume Data (end)
\section{Guards in the Backend}
\label{sec:Guards in the Backend}
